Manuscript title here

### 2024-09-10 manuscript.Rmd
råka fixa till referenserna
tog tillbaka det jag råka tabort

---
title: "Manuscript title here"
output: html_document
author: "Your name here"
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

1  Varför trauma är viktigt
    Mortalitet för unga
    Varför chock patienter är viktig subdel av akut omhändertagande av trauma pat
      Kanske (infoga källa) att chock är stor del av mortalitet.

Trauma is a major global public health concern in form of death and disability. It affects over four milion peole each year, among those younger and economicaly active(??) populations are the most affected(källa). The initial management of trauma patients once they reach hospital is very time sensitive and error prone(källa), especially in patients arriving in shock(källa). Therefore, careful studying of this part is of importance.

@@

Shock are generally defined as circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[]. There are five types of causes, which are hypovolemia, obstructive, neurogenic, cardiogenic and distributive. These causes can occur alone or in combination. The most usuall one during trauma are hypovolemia in form of a large bleeding.

@@

Clinically to be able to differentiate the degree of shock, usually the ATLS classification for major haemorrhage are used. The classification are divided into four classes, based on the patients: estimated bloodloss, heart rate, bloodpressure, pulse pressure, respiratory rate, mental status and urine output. However this only suits hypovolemic shocks and not the other types.

4  Kliniskt för att förbättra omhändertagande används OFI
    Förklara vad OFI är för något

To be able to asses and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI). OFI is multidiciplinary morbidity and mortality reviews, where oppertunity for improvemetns are identified in the stracture and clinical process of trauma care(Hussein).

@@

```{r main, echo=FALSE}
source("main.R") ## This "imports" the main script file of your project and run any code in it
```

>>>>

---
title: "Manuscript title here"
author: "Your name here"
output:
  word_document: default
  html_document: default
bibliography: bibliography.bib
csl: vancouver.csl
---

@@

1  Varför trauma är viktigt
    Mortalitet för unga
    Varför chock patienter är viktig subdel av akut omhändertagande av trauma pat
      Kanske att chock är stor del av mortalitet.

Trauma is a major global public health concern in form of death and disability. It affects over four milion peole each year, among those younger populations are the most affected[@noauthor_injuries_nodate]. The initial management of trauma patients once they reach hospital is very time sensitive and error prone[@mackersie_pitfalls_2010]. This combined with the fact that trauma patients in shock have a higher mortality rate, than those not[@vang_shock_2022], makes it important to study and improve the care for trauma patients in shock.

@@

Shock are generally defined as circulatory failure leading to insuficient perfusion and oxygenation of tissue and organ[@leech_shock_2023]. There are five types of causes, which are hypovolemia, obstructive, neurogenic, cardiogenic and distributive. These causes can occur alone or in combination. The most usuall one during trauma are hypovolemia in form of a large bleeding[@teixeira_preventable_2007].

@@

Clinically to be able to differentiate the degree of shock, the ATLS classification for major haemorrhage can be used. The classification are divided into four classes, based on the patients: estimated bloodloss, heart rate, bloodpressure, pulse pressure, respiratory rate, mental status and urine output. However this only suits hypovolemic shocks and not the other types.

Skriva om stycke 4 och istället lägga in innehåll om använda retts som en fortsättnig på stycke 3.

  (4  Kliniskt för att förbättra omhändertagande används OFI
    Förklara vad OFI är för något

  To be able to asses and improve the quality of trauma care, improvement programmes aim to identify opportunities for improvements(OFI). OFI is multidiciplinary morbidity and mortality reviews, where oppertunity for improvemetns are identified in the structure and clinical process of trauma care(Hussein).)

@@
